Making multiple-locus association mapping on a genome-wide scale routine

### 2017-10-31 AM+draft.tex
31Oct Still working on methods - glmnet

\subsubsection{Reducing the dimension of the model}
In practice, estimating the parameters of (\ref{eq2}) can be demanding, computationally. 
The vector $a_{-S}$ has $L-s$ random effects where in modern genome-wide association studies, 
$L$, the number of snp, can be extremely large.  An alternative model is given by 
Verbyla. 
They show how to reformulate (\ref{eq2}) to be a model with a random effect with only $n$ elements
\begin{equation}
\label{eq3}
u_g = \sum_{k=1}^s  m_{S_k} a_{S_k} + (M_{-S} M_{-S}^T)^{1/2} a^*_{-S}
\end{equation}
where $a^* \sim N(0, \sigma_a^2 I^{(n \; \times \;  n)})$, and 
$(M_{-S} M_{-S}^T)^{1/2}$ can be calculated via single value decomposition (ref).  
Although it may not be obvious, the two models are equivalent, 
having identical variance structures. Yet, the computational cost of model (\ref{eq3}) compared to 
model (\ref{eq2}) is much less due to the random term having only $n$ effects needing estimating.

Verbyla go on to show how to recover $\widetilde{a}$ from estimates from model (3) with 
\begin{equation}
\widetilde{a} = \left [ M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \right ] \widetilde{a}^*
\end{equation}
where its variance matrix is
\begin{equation}
\label{eq4}
\textrm{var}(\widetilde{a}) = M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \textrm{var}(\widetilde{a}^*) (M_{-S} M_{-S}^T)^{-1/2} M_{-S}
\end{equation}
These values are needed in order to calculate the score statistic $t_j^2$ for identifying the snp in strongest association with the trait. 
Fortunately, when calculating $t_j^2$, only the diagonal elements of the variance matrix are needed which simplifies the  calculation. 
of (\ref{eq4}).

>>>>

\subsubsection{Reducing the dimension of the model}
In practice, estimating the parameters of (\ref{eq2}) can be demanding, computationally. 
The vector $a_{-S}$ has $L-s$ random effects where in modern genome-wide association studies, 
$L$, the number of snp, can be extremely large.  An alternative model is given by 
Verbyla \cite{verbyla2012rwgaim,verbyla2014whole}. 
They show how to reformulate (\ref{eq2}) to be a model with a random effect with only $n$ elements
\begin{equation}
\label{eq3}
u_g = \sum_{k=1}^s  m_{S_k} a_{S_k} + (M_{-S} M_{-S}^T)^{1/2} a^*_{-S}
\end{equation}
where $a^* \sim N(0, \sigma_a^2 I^{(n \; \times \;  n)})$, and 
$(M_{-S} M_{-S}^T)^{1/2}$ can be calculated via single value decomposition (ref).  
Although it may not be obvious, the two models are equivalent, 
having identical variance structures. Yet, the computational cost of model (\ref{eq3}) compared to 
model (\ref{eq2}) is much less due to the random term having only $n$ effects needing estimating.

Verbyla \cite{verbyla2012rwgaim,verbyla2014whole} go on to show how to recover $\widetilde{a}$ from estimates from model (3) with 
\begin{equation}
\widetilde{a} = \left [ M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \right ] \widetilde{a}^*
\end{equation}
where its variance matrix is
\begin{equation}
\label{eq4}
\textrm{var}(\widetilde{a}) = M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \textrm{var}(\widetilde{a}^*) (M_{-S} M_{-S}^T)^{-1/2} M_{-S}
\end{equation}
These values are needed in order to calculate the score statistic $t_j^2$ for identifying the snp in strongest association with the trait. 
Fortunately, when calculating $t_j^2$, only the diagonal elements of the variance matrix are needed which simplifies the  calculation. 
of (\ref{eq4}).

\subsection{Comparison Methods}

We compared the performance of Eagle against seven other association mapping applications. 
Five of these applications, BigRR, glmnet, FaST-LASSO, MMLM, r2VIM implement methods that have been applied to the multiple locus analysis of data from genome-wide association studies. 
The other two applications, GEMMA, and FaST-LMM implement single-locus association mapping methods. 
The packages 
differ in their features (summarized in Table X).  The packages also differ in their underlying methods (summarized below).

\subsubsection{Multiple locus methods}

\textbf{BigRR:}  In the BigRR package,  generalised ridge regression is implemented  for
the analysis of genetic data \cite{shen2013novel}.  Generalised ridge regression 
 is a regularisation method in that it produces regression solutions 
that have higher bias than ordinary least squares but reduced variance for improved prediction accuracy. The underlying 
model is  $y = X \beta + M u + e$ where $u^{(L \times 1)}$ is a vector of genetic effects, and each genetic 
effect is normally distributed with its own variance such that $u_l \sim N(0, \sigma^2_l); l=1,2, \ldots, L$.  The 
amount of shrinkage is allowed to vary across the genetic (snp) effects. This differs to 
standard ridge regression where the shrinkage is constant. 
Generalised ridge regression is apt at estimating the effect sizes of snp on a genome-wide scale but  obtaining their 
statistical significance analytically is difficult. Shen \cite{shen2013novel} suggested permutation as a way of calculating significance 
empirically. We instead opted for stability selection (see below of details).

\textbf{glmnet:}  Glmnet is an R package where elastic net regression is implemented. Elastic net is a regularisation method that 
has been applied to generalised linear models \cite{zou2005regularization}.

Glmnet is a general purpose R package for statistical analysis. It implements elastic net regression, 
a regularisation method that has been applied to generalised linear models \cite{zou2005regularization}.   
Glmnet was not specifically built for association mapping. However, we included it amongst the applications to benchmark 
Eagle against because of the interest surrounding its use for analysing data from genetic
studies \cite{cho2010joint,huang2015empirical}.  The regression model we used for analysis is 
$y = X \beta + \sum_{l=1}^L m_l \theta_l + e$ where $\theta_l; l=(1,2, \ldots, L)$ is the coefficient for the $l$th snp. 
Elastic net allows us to obtain parameter estimates from this model via a penalty function that is a mixture of 
$\ell_1$ (lasso) and $\ell_2$ (ridge regression).

Although glmnet has not been specifically built for association mapping, interest surrounds its use for the analysis of data from 
an association study.  \cite{cho2010joint,huang2015empirical}. 
The analysis model for association mapping 
$y = X \beta + \sum_{l=1}^L m_l \theta_l + e$ where $\theta_l; l=(1,2, \ldots, L)$ is the regression coefficient for the $l$th snp. 
It's penalty function for regularisation is a mixture of $\ell_1$ (lasso) and $\ell_2$ (ridge regression).  
This allows elastic net to avoid some of the
limitations of lasso (ref) while still being able to perform shrinkage and variable selection, simultaneously.  The glmnet package 
implements a modification of the LARS algorithm \cite{efron2004least} for finding the optimal regularisation (solution) path. However, 
from exploratory analyses of the simulated data, we found this algorithm yielded a high proportion of false positives. 
We instead fixed the shrinkage parameter to a suitable value 
and employed stability selection to estimate the significance of the snp regression coefficients, empirically (see below).

  MMLM

  r2VIM

  LMM-lasso

Single locus methods

gemma

Fast-LMM
